Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2021-08-24 Frontiers/Frontiers.tex
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic in the Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of United States Midwest. In general, results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). At first cohort, Palmer amaranth produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas Palmer amaranth produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} at second cohort in bareground, soybean and corn, respectively. The plant height was more impacted by Palmer amaranth growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increase as biomass and height increases. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitat is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability establish into cropping systems, including crop rotation (not only corn and soybean), early/late crop planting, row spacing, and cover crops. Our results also shed a light on sex dimorphism as a future Palmer amaranth ecological management.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and reliance on herbicides for weed management are the main
human-mediated selection of Palmer amaranth into cropping systems (Ward
et al., 2013). On the other hand, Palmer amaranth is a prolific seed
producer with a C4 photosynthetic apparatus (Wang et al., 1992). With a
dioecy nature, Palmer amaranth male and female plants are obligate
outcrosser species, increasing the chances of exchanging adaptive traits
among plants (Jhala et al., 2021; Oliveira et al., 2018). Also, Palmer
amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage
systems (Price et al., 2011), and spread across locations through farm
equipment (Sauer, 1972), manure (Hartzler and Anderson, 2016), animals
(Farmer et al., 2017). The dispersal capacity of Palmer amaranth makes
the species one of the most successful cases of weed adaptation to
current cropping systems.

@@

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, Palmer amaranth range of expansion into new habitats
is increasing. It seems fated the need to manage Palmer amaranth in
agronomic crops throughout multiple environments in the near future.
Strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built on
minimizing the species ability to adapt, grow and develop into
agroecossystems.

@@

A Palmer amaranth accession (Per1) from Perkins County, Nebraska. with
no reported herbicide resistance according to Oliveira et al. (2021) was
selected for this study. Three weeks prior to the onset of the field
experiments, seeds were planted in plastic trays containing potting-mix.
Emerged seedlings (1 cm) were transplanted into 200
cm\textsuperscript{-3} plastic pots (a plant pot\textsuperscript{-1}).
Palmer amaranth seedlings were supplied with adequate water and kept
under greenhouse conditions at the University of Wisconsin-Madison,
University of Nebraska-Lincoln, and Western Illinois University; and
kept outdoors at the Perkins extension office in Grant, NE until the 2-3
leaf stage (5 to 8 cm height) when they were transported to the field.

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE
(40.57'N, 9814'W), Grant, NE (('N, 'W)), Lincoln, NE ((41.16'N,
96.42'W)), and Macomb, IL (('N, 'W)).

@@

 & Hybrid &  & DKC60-67 &  &  & \\

 & Seeding rate &  & 86487 &  &  & \\

 & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

 & Stage 1 cohort & V2-3 &  &  &  & \\

\multirow{-5}{*}{\raggedright\arraybackslash } & Stage 2 cohort & V6-7 &  &  &  & \\

@@

 & Stage 1 cohort & V1-2 &  &  &  & \\

 & Stage 2 cohort & V5-6 &  &  &  & \\

\multirow{-5}{*}{\raggedright\arraybackslash } & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

Planting day &  & May 10 to 20 & May 10 to 14 &  &  & \\

@@

 & Type &  & Crete Silt Loam &  &  & \\

 & Ratio (sand-clay-silt) &  & 58-25-6.5 &  &  & \\

@@

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) &  & 3 &  &  & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} glyphoste, 840 g ae ha \textsuperscript{b} S-metolachor, g ai ha \textsuperscript{c} S-metolachlor + trazine + mesotrione, + bicyclopyrone, g ai ha \textsuperscript{d} saflufenacil + imazethapyr + pyroxasulfone, g ai ha}\\
\end{tabular}}
\end{table}

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplanting timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop (Figure 2). After a week, one plant was eliminated and one was
kept, resulting in 24 plants per experimental unit and transplanting
time (Figure 2). When needed, Palmer amaranth plants were supplied with
water during the first week after transplanting.

@@

The statistical analyses were performed using R statistical software
version 4.0.1. Data analyses are stored at Zenodo (Oliveira, 2021).

Analyses of Palmer amaranth height and biomass were performed with a
linear mixed model using \emph{lmer} function from ``lme4'' package
(Bates et al., 2015). Plant height and biomass were log transformed to
meet model assumption of normality. In the model, crop (bareground,
corn, soybean) was the fixed effect and year nested with location the
random effects. Analysis of variance was performed with \emph{anova}
function from ``car'' package (Fox and Weisberg, 2018). Marginal means
and compact letter display were estimated with \emph{emmeans} and
\emph{cld} from packages ``emmeans'' and ``multcomp'' (Hothorn et al.,
2008).

The Palmer amaranth flowering timing was estimated as cumulative
flowering across all location. Dataset was nested but id did not include
Grant, NE. Palmer amaranth cumulative flowering estimation was
determined using an asymmetrical three parameter log logistic Weibull
model of the drc package (Ritz et al., 2015).

@@

Palmer amaranth sex dimorphism was fitted to a binary logistic
regression (Bangdiwala, 2018). Binary logistic regression is used for
predicting binary classes, such as the probability of a plant being
female in a dioecious species. Prior to the analysis, all missing values
were removed from the dataset. Also, data from Grant was not used in
this analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking how the model generalized on unseen dataset. With 80\% dataset,
a binary response variable, male (0) and female (1), was fitted to a
generalized linear model (\emph{glm} function) including day of year
harvest, height, weight, crop and month as independent variables
(without interaction). The model family was binomial with a logit
function. The model fit was assessed through pseudo R-squared values
(McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using
\emph{nagelkerke} function (``rcompanion'' package). The marginal
effects computation was performed with Average Marginal Effects (AMEs)
at every observed value of x and average across the results (Leeper,
2017) using \emph{margins} function from ``margins'' package. The rest
20\% dataset was predicted using \emph{predict} function with a cutoff
estimation for male or female using \emph{performance} function. The
model quality prediction from the classification algorithm was measured
with precision (\emph{precision} function), recall (\emph{recall}
function) and F1-score (\emph{f\_meas} function) using the ``yardstick''
package. The precision determines the accuracy of positive predictions
(female plants), recall determines the fraction of positives that were
correctly identified, and F1-score is a weighted harmonic mean of
precision and recall with the best score of 1 and the worst score of 0.
F1-score conveys the balance between the precision and the recall
(Yacouby and Axman, 2020).

@@

The model fit was 0.23, 0.32, 0.40 using pseudo R-squared test from
McFadden, Cox and Snell, and Cragg and Uhler, respectively. The
likelihood ratio test showed a p-value of \textless{} 0.00. The average
marginal effects showed that Palmer amaranth growing in corn resulted in
14.8\% less females plants (Table 2). Moreover, increasing a unit doy
increases the probability of having a female plant by 0.4\% (Table 2 and
Figure4A). Similar trend is observed for weight as well as height,
whereas the probability of being female increase by 0.2\% (Figure 4B)
and 0.1\% (Figure 4C) when a unit of weight (g) and height (cm)
increases, respectively.

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal means of sex dimorphism logistic model. Factor pararemter values (e.g. crop) is shown related to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

The model accuracy evaluation accuracy in the 20\% test dataset was 0.62
with a cutoff value for female and male plants of 0.43. The model
classification showed a precision of 0.64, recall of 0.66, and a F means
score of 0.65. In addition, the area under the curve was 0.64.

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern
and sex varied within crops and cohort timings. In general, first cohort
of Palmer amaranth plants were heavier and taller when compared to the
second cohort. At first cohort, resources (e.g., soil nutrients) and
conditions (e.g., light) were more timely available for the species.
High biomass and taller Palmer amaranth plants are likely a weed
strategy to compete for light in between crop rows in absence of canopy.
In such conditions, Palmer amaranth showed an extraordinary plasticity
to adapt upon the agroecosystem. This is evident when comparing Palmer
amaranth canopy shape, and its extended flowering pattern when growing
into corn compared to soybean. The Palmer amaranth competition (e.g.,
light) strategy was to mimic the crop grow and development (Figure 6).
These results suggests that Palmer amaranth can fast evolve life-history
traits to adapt into cropping systems and cultural practices, which was
also showed in a study varying nitrogen fertilization (Bravo et al.,
2018). Our results highlight the Palmer amaranth as a threat to field
crops as breeding more competitive crop varieties is likely to select
more competitive biotypes (Bravo et al., 2017).

Palmer amaranth grow and development in second cohort was limited due to
the crop competitive ability at advanced development stages. Palmer
amaranth was transplanted when corn canopy was nearly closed, which
reduced Palmer amaranth competitiveness. As a result, Palmer amaranth
height and biomass was lower compared to its first cohort. Under crop
canopy (e.g., second cohort), Palmer amaranth flowering window was near
to similar in corn and soybean. Palmer amaranth growing without crop
competition produced the highest amounts of biomass and less extended
flowering window. The Palmer amaranth strategy in bareground was to
invest biomass in growing plant width and height. Nonetheless, Palmer
amaranth produced 21\% less biomass in second cohort compared to first
cohort timing. In a bareground study, early emerged Palmer amaranth
without competition was 50\% taller than late emerged plants (Webster
and Grey, 2015). These results suggest that crop competition is not the
only factor limiting late Palmer amaranth establishment. The limited
growth of Palmer amaranth at second cohort is likely a reduced plant
response to day length, light availability and thermal units (e.g,
growing degree days). It is hypothesize that reduced day length
contributed to smaller plants at second cohort as well as shorter
flowering period. A study in North Carolina and Illinois predicted that
less than 10\% Palmer amaranth seedlings emergence occurred after June
(Piskackova et al., 2021). In addition, Palmer amaranth negative impact
on soybean (Korres et al., 2020) and cotton (Webster and Grey, 2015)
yields was higher when plants were established near to crop planting.

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, Palmer
amaranth growing at first cohort accumulated an overall 42\% more
biomass when compared to second cohort. Therefore, Palmer amaranth
plants growing in the second cohort is likely to produce less seeds
regardless the cropping system. Our observation is consistent with the
findings that first Palmer amaranth cohort produced 50\% more seeds per
plant than Palmer amaranth plants established six weeks later in
bareground (Webster and Grey, 2015). Still, seed production at second
cohort is likely to replenish the soil seedbank. Seed production and
deposition in the seedbank is also a key factor for species perpetuation
(Menges, 1987). Palmer amaranth can produce hundred thousands seeds per
plant (Schwartz et al., 2016; Keeley et al., 1987), and stay viable
buried in the seedbank for at least 36 months (Sosnoskie et al., 2013).
Therefore, preventing Palmer amaranth seed production or/and seed
migration to its non-native habitat is an essential strategy to minimize
weed impact into agroecossystem (Davis et al., 2015).

An ecological approach to reduce seed production in Palmer amaranth is
understanding its flowering window. Our study suggests that Palmer
amaranth flowering was slightly influenced by cropping systems and
cohort timings. Palmer amaranth growing in bareground and corn resulted
in the overall shortest and longest flowering window, respectively. When
growing in soybean, Palmer amaranth flowering window was similar to
bareground at first cohort but similar to corn at second cohort timing.
Plant flowering initiation is complex and depends on biological and
ecological factors (Lang, 1965). We hypothesize that when growing in
high competition (e.g., second cohort) and/or under low light/short day
length, Palmer amaranth plants tend to initiate flowering early, as well
as having an extended flowering window. Early flower initiation is plant
strategy when growing in stress conditions. For example, when growing
under water stress, early flowering in Palmer amaranth resulted in a
mismatch between female and male plants by seven days (Mesgaran et al.,
2021). A mismatch in Palmer amaranth male and female flowering period
can minimize plant outcross, and thus reduce plant seed production and
exchange of resistant alleles (Jhala et al., 2021). Sex dimorphism
manipulation is considered a potential ecological pest control
(McFarlane et al., 2018; Schliekelman et al., 2005).

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of United States Midwest. In general, results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). At first cohort, Palmer amaranth produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas Palmer amaranth produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} at second cohort in bareground, soybean and corn, respectively. The plant height was more impacted by Palmer amaranth growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increase as biomass and height increases. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from United States Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability establish into cropping systems, including crop rotation (not only corn and soybean), early/late crop planting, row spacing, and cover crops. Our results also shed a light on sex dimorphism as a future Palmer amaranth ecological management.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selection of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecy nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth small seed size (e.g, 1 mm) tend to thrive
in no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), manure (Hartzler and Anderson,
2016), animals (Farmer et al., 2017). The dispersal capacity of Palmer
amaranth makes the species one of the most successful cases of weed
adaptation to current cropping systems.

@@

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Waterhemp, an
\emph{Amaranthus} species, is the most troublesome species in that
geography (Tranel et al., 2011). Nonetheless, invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, Palmer amaranth range of expansion into new habitats
can increase. It seems certain the need to manage Palmer amaranth in
agronomic crops throughout novel environments in the near future.
Strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built on
minimizing the species ability to adapt, grow and develop into
agroecossystems.

@@

A Palmer amaranth accession (Kei3) from Perkins County, Nebraska. with
no reported herbicide resistance according to Oliveira et al. (2021) was
selected for this study. Three weeks prior to the onset of the field
experiments, seeds were planted in plastic trays containing potting-mix.
Emerged seedlings (1 cm) were transplanted into 200
cm\textsuperscript{-3} plastic pots (a plant pot\textsuperscript{-1}).
Palmer amaranth seedlings were supplied with adequate water and kept
under greenhouse conditions at the University of Wisconsin-Madison,
University of Nebraska-Lincoln, and Western Illinois University; and
kept outdoors at the Perkins extension office in Grant, NE until the 2-3
leaf stage (5 to 8 cm height) when they were transported to the field.

@@

The experiment was conducted in 2018 and 2019 under field conditions at
five locations: Arlington, WI (43°18'N, 89°29'W), Clay Center, NE
(40.57'N, 9814'W), Grant, NE (XXX'N, XXX'W), Lincoln, NE (41.16'N,
96.42'W), and Macomb, IL (XXX'N, XXX'W).

@@

 & Hybrid & NK0142 3120-EZ1 & DKC60-67 &  &  & \\

 & Seeding rate & 88956 & 86487 &  &  & \\

 & Weed control & glyphosate\textsuperscript{a} / \emph{S}-metolachor\textsuperscript{b} & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{c} &  &  & \\

 & Stage at 1\textsuperscript{st} cohort & V2-3 &  &  &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V6-7 &  &  &  & \\

 & Planting day & April 30, 2018 / May 5, 2019 & May 10, 2018/19 &  &  & \\

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha\textsuperscript{-1} &  &  &  & \\

@@

 & Weed control & glyphosate / \emph{S}-metolachor & saflufenacil + imazethapyr\textsuperscript{d} + pyroxasulfone &  &  & \\

 & Stage at 1\textsuperscript{st} cohort & V1-2 &  &  &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V5-6 &  &  &  & \\

\multirow{-6}{*}{\raggedright\arraybackslash } & Planting day & May 5, 2018 / May 10, 2009 & May 14, 2018/19 &  &  & \\

@@

 & Type & Plano-silt-loam & Crete Silt Loam &  &  & \\

 & Ratio (sand-clay-silt) & 10-64-26 & 17-58-25 &  &  & \\

@@

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 &  &  & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} glyphoste, 840 g ae ha \textsuperscript{b} S-metolachor, 1324 g ai ha; \textsuperscript{c} S-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha; \textsuperscript{d} saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha}\\
\end{tabular}}
\end{table}

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplanting timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop. After a week, one plant was eliminated and one was kept, resulting
in 24 plants per experimental unit and transplanting time (Figure 2).
When needed, Palmer amaranth plants were supplied with water during the
first week after transplanting.

@@

The statistical analyses were performed using R statistical software
version 4.0.1 (Team, 2021). Data analyses are stored at Zenodo
(Oliveira, 2021).

Analyses of Palmer amaranth height and biomass were performed with a
linear mixed model using \emph{lmer} function from ``lme4'' package
(Bates et al., 2015). Plant height and biomass were log transformed to
meet model assumption of normality. In the model, crop (bareground,
corn, soybean) and cohort time (first and second) were the fixed effects
and year nested with location the random effects. Analysis of variance
was performed with \emph{anova} function from ``car'' package (Fox and
Weisberg, 2018). Marginal means and compact letter display were
estimated with \emph{emmeans} and \emph{cld} from packages ``emmeans''
(Lenth et al., 2021) and ``multcomp'' (Hothorn et al., 2008),
respectively.

The Palmer amaranth flowering timing was estimated as cumulative
flowering across all location. Dataset was nested but it did not include
Grant, NE. Palmer amaranth cumulative flowering estimation was
determined using an asymmetrical three parameter log logistic Weibull
model of the drc package (Ritz et al., 2015).

@@

Palmer amaranth sex dimorphism was fitted to a binary logistic
regression (Bangdiwala, 2018). Binary logistic regression is used for
predicting binary classes, such as the probability of a plant being
female in a dioecious species. Prior to the analysis, all missing values
were removed from the dataset. Also, data from Grant was not used in
this analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking model performance on unseen dataset. With 80\% dataset, a
binary response variable, male (0) and female (1), was fitted to a
generalized linear model (base R \emph{glm} function) including day of
year harvest, height, weight, crop and month as independent variables
(without interaction). The model family was binomial with a logit
function. The model fit was assessed through pseudo R-squared values
(McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using
\emph{nagelkerke} function from ``rcompanion'' package (Mangiafico,
2021). The marginal effects computation was performed with Average
Marginal Effects (AMEs) at every observed value of x and average across
the results (Leeper, 2017) using \emph{margins} function from
``margins'' package (Leeper et al., 2021). The rest 20\% dataset was
predicted using \emph{predict} function with a cutoff estimation for
male or female using \emph{performance} function. The model quality
prediction from the classification algorithm was measured with precision
(\emph{precision} function), recall (\emph{recall} function) and
F1-score (\emph{f\_meas} function) using the ``yardstick'' package (Kuhn
et al., 2021). The precision determines the accuracy of positive
predictions (female plants), recall determines the fraction of positives
that were correctly identified, and F1-score is a weighted harmonic mean
of precision and recall with the best score of 1 and the worst score of
0. F1-score conveys the balance between the precision and the recall
(Yacouby and Axman, 2020).

@@

The model fit was 0.23, 0.32, 0.40 using pseudo R-squared test from
McFadden, Cox and Snell, and Cragg and Uhler, respectively. The
likelihood ratio test showed a p-value of \textless{} 0.00. The average
marginal effects showed that Palmer amaranth growing in corn resulted in
14.8\% less females plants (Table 2). Moreover, increasing a unit doy
increases the probability of having a female plant by 0.4\% (Table 2 and
Figure4A). Similar trend is observed for weight (Figure 4B) as well as
height (Figure 4C), whereas the probability of being female increase by
0.2\% and 0.1\% when a unit of weight (g) and height (cm) increases,
respectively.

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal means of Palmer amaranth sex dimorphism logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME\textsuperscript{a} & SE\textsuperscript{b} & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

The model accuracy evaluation accuracy in the 20\% test dataset was 0.62
with a cutoff value for female and male plants of 0.43. The model
classification showed a precision of 0.64, recall of 0.66, and a
F1-score of 0.65. In addition, the area under the curve was 0.64.

@@

Our study showed that Palmer amaranth biomass, height, flowering window
and sex varied within crops and cohort timings. In general, first cohort
of Palmer amaranth plants were heavier and taller when compared to the
second cohort. At first cohort, resources (e.g., soil nutrients) and
conditions (e.g., light) were more timely available for the species.
High biomass and taller Palmer amaranth plants are likely a weed
strategy to compete for light in between crop rows in absence of canopy.
In such conditions, Palmer amaranth showed an extraordinary plasticity
to adapt upon the agroecosystem. This is evident when comparing Palmer
amaranth canopy shape. The Palmer amaranth competition (e.g., light)
strategy was to mimic the crop grow and development (Figure 6). These
results suggests that Palmer amaranth can fast evolve life-history
traits to adapt into cropping systems and cultural practices, which was
also showed in a study varying nitrogen fertilization (Bravo et al.,
2018). Our results highlight the Palmer amaranth as a threat to field
crops as breeding more competitive crop varieties is likely to select
more competitive weed biotypes (Bravo et al., 2017).

Palmer amaranth grow and development in second cohort was limited due to
the crop competitive ability at advanced development stages. Plants were
transplanted at greater crop height, which reduced Palmer amaranth
competitiveness. As a result, Palmer amaranth height and biomass was
lower compared to its first cohort. Palmer amaranth growing without crop
competition produced the highest amounts of biomass. The Palmer amaranth
strategy in bareground was to invest biomass in growing plant width and
height. Nonetheless, Palmer amaranth produced 17\% less biomass in
second cohort compared to first cohort timing. In a bareground study,
early emerged Palmer amaranth without competition was 50\% taller than
late emerged plants (Webster and Grey, 2015). These results suggest that
crop competition is not the only factor limiting late Palmer amaranth
establishment. The limited growth of Palmer amaranth at second cohort is
likely a reduced plant response to day length, light availability and
thermal units (e.g, growing degree days). The \emph{Amaranthus} species
are sensitive to photoperiod (Wu and Owen, 2014). It is hypothesize that
reduced day length contributed to smaller plants at second cohort
regardless the crop. A study in North Carolina and Illinois predicted
that less than 10\% Palmer amaranth seedlings emergence occurred after
June (Piskackova et al., 2021). In addition, Palmer amaranth negative
impact on soybean (Korres et al., 2020) and cotton (Webster and Grey,
2015) yields was higher when plants were established near to crop
planting.

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, plants
growing at first cohort accumulated an overall 42\% more biomass when
compared to second cohort. Therefore, Palmer amaranth plants growing in
the second cohort is likely to produce less seeds regardless the crop.
Our observation is consistent with the findings that first Palmer
amaranth cohort produced 50\% more seeds per plant than Palmer amaranth
plants established six weeks later in bareground (Webster and Grey,
2015). Still, seed production at second cohort is likely to replenish
the soil seedbank. Seed production and deposition in the seedbank is
also a key factor for species perpetuation (Menges, 1987). Palmer
amaranth can produce hundred thousands seeds per plant (Schwartz et al.,
2016; Keeley et al., 1987), and stay viable buried in the seedbank for
at least 36 months (Sosnoskie et al., 2013). Therefore, preventing
Palmer amaranth seed production or/and seed migration to its non-native
habitat is an essential strategy to minimize the species impact into
agroecossystem (Davis et al., 2015).

An ecological approach to reduce seed production in Palmer amaranth is
understanding its flowering window. Our study suggests that Palmer
amaranth flowering was slightly influenced by crops and cohort timings.
Palmer amaranth growing in bareground and corn resulted in the overall
shortest and longest flowering window, respectively. The shortest
flowering window of second cohorts of Palmer amaranth growing in
bareground highlighted the impact of photoperiod on flowering. When
growing in soybean, Palmer amaranth flowering window was similar to
bareground at first cohort but similar to corn at second cohort timing.
Plant flowering initiation is complex and depends on biological and
ecological factors (Lang, 1965). We hypothesize that when growing in
high competition (e.g., second cohort), Palmer amaranth plants tend to
initiate flowering early, as well as having an extended flowering
window. A study has shown that non-native Palmer amaranth initiated
flowering two weeks prior to the native waterhemp in Iowa (Baker, 2021).
Early flower initiation could be also a plant strategy when growing in
stress conditions. For example, when growing under water stress, early
flowering in Palmer amaranth resulted in a mismatch between female and
male plants by seven days (Mesgaran et al., 2021). A mismatch in Palmer
amaranth male and female flowering period can minimize plant outcross,
reducing plant seed production and/or exchange of resistant alleles
(Jhala et al., 2021). Sex dimorphism manipulation is considered a
potential ecological pest control (McFarlane et al., 2018; Schliekelman
et al., 2005).
